---
title: "Comparison of accuracies of RapidEye and Landsat 8 OLI images in mapping forest attributes"
author: "Oyedayo Oyelowo"
date: "9 April 2017"
output: html_document
---


# MAP TREE ATTRIBUTES USING RAPIDEYE AND LANDSAT 8 IMAGES AND kNN METHOD
```{r, message=FALSE, warning=FALSE, echo=FALSE}
##########################
#SET WORKING DIRECTORY
##########################

# clean the working environment and graphics
rm(list=ls()); graphics.off()

# specify path to your working directory
setwd("C:/Users/oyelo/Desktop/advanced remote sensing/exercise 3/BIODEV training manual, data and scripts-20170403/R/R/Data")
getwd()

#Set path for packages installation and load the packages
.libPaths("C:/Users/oyelo/Desktop/advanced remote sensing/exercise 3")

library(sp)
library(rgdal)
library(raster)
library(yaImpute)
library(fastICA)



##########################
#RUN FUNCTIONS
##########################

#####
# Function for accuracy assessment
#####

print.accuracy <- function(obs, fit, main) {
  print(main)
  res <- fit-obs
  bias <- signif(mean(res), 3); cat("BIAS:", bias)
  bias.r <- signif(bias/mean(obs)*100, 3); cat( " (",bias.r,"%)\n", sep="")
  RMSE <- signif(sqrt(sum((res^2)/length(res))), 3); cat("RMSE:",RMSE)
  RMSE.r <- signif(RMSE/mean(obs)*100, 3); cat( " (",RMSE.r,"%)\n", sep="")
  R2 <- signif(cor(obs,fit,method="pearson")^2, 3); #pseudo R^2
  cat("Pseudo R.squared: ", R2,"\n\n", sep="")
  ret <- c(bias, bias.r, RMSE, RMSE.r, R2)
}


#####
# Function for leave-one-out cross-validation
#####

cv.plot <- function(data, yvar, xvars, k, yai.method) {
  pred <- numeric(nrow(data)) #empty vector for results
  for(i in 1:nrow(data)){
    ref <- which(row.names(data)!=i)
    val <- which(row.names(data)==i)
    x <- as.data.frame(data[,xvars])
    y <- as.data.frame(data[ref, yvar])
    names(y) <- yvar
    row.names(y) <- row.names(data)[-val]
    knn <- yai(x=x, y=y, k=k, data=data, method=yai.method)
    imp <- impute(knn, k=k, method="dstWeighted", ancillaryData=y)
    pred[val] <- imp[as.numeric(row.names(imp)) %in% val, yvar]
  }
  
  #print results
  res <- print.accuracy(data[,yvar], pred, yvar)
  xlim <- max(data[,yvar])
  plot(data[,yvar], pred, main=yvar, xlab="Observed", ylab="Predicted",
       xlim=c(0,xlim), ylim=c(0,xlim))
  abline(0,1)
  return(res)
}

##########################
#LOAD INPUT DATA 
##########################

# load required field data
load("Kou_input_plots.rda")   #"Plots" are required because of coordinates
load("PlotData_Kou.rda")      #Plot-level results
```


##########################
#RAPIDEYE IMAGE
##########################'
```{r, warning=FALSE}

# "re_refl.tiff" is a spatial subset of Rapid Eye satellite image

# load multiband satellite image stack
rapideye<- stack("C:/Users/oyelo/Desktop/advanced remote sensing/exercise 3/BIODEV training manual, data and scripts-20170403/R/R/Data/re_refl.tif")
rapideye#see properties of RasterStack object

# aggregare ERapid EYe
rapideye.30m<-aggregate(rapideye, fact=6, fun=mean)
rapideye.30m

#remove unnecessary files
rm(rapideye)

# the image has five spectral bands corresponding to blue, green, red, red edge
# near infrared (NIR)  parts of the electromagnetic
# spectrum

# rename bands
bands_re <- c("b.re","g.re","r.re","rededge","nir.re")
names(rapideye.30m) <- bands_re


par(mfrow=c(1,2))
# visualize the image
plotRGB(rapideye.30m, r="r.re", g="g.re", b="b.re", stretch="lin", main="RE truecolour")

# this is so called "true-color" composite as it best corresponds to colors seen by eye

# several "false-color" composites exist for visualization
# here, green vegetation is seen in red (strong reflectance in near infrared)
plotRGB(rapideye.30m, r="nir.re", g="r.re", b="g.re", stretch="lin", main="RE false colour")
```



##########################
#LANDSAT IMAGE
##########################
```{r, warning=FALSE}

# "Landsat.tiff" is a spatial subset of Landsat 8 OLI satellite image
# acquired 11 October 2013

# load multiband satellite image stack
landsat<- stack("C:/Users/oyelo/Desktop/advanced remote sensing/exercise 3/BIODEV training manual, data and scripts-20170403/R/R/Data/ls8_sr_2014_02_16.tif")
landsat#see properties of RasterStack object

# the image has six spectral bands corresponding to blue, green, red, 
# near infrared (NIR) and short waveinfrared (SWIR) parts of the electromagnetic
# spectrum

# rename bands
bands_l8 <- c("b","g","r","nir","swir1","swir2")
names(landsat) <- bands_l8


par(mfrow=c(1,2))
# visualize the image
plotRGB(landsat, r="r", g="g", b="b", stretch="lin", main="LS8 true colour")
# the image covers 12 km x 12 km area around sentinel site mid-point
# this is so called "true-color" composite as it best corresponds to colors seen by eye

# several "false-color" composites exist for visualization
# here, green vegetation is seen in red (strong reflectance in near infrared)
plotRGB(landsat, r="nir", g="r", b="g", stretch="lin", main="LS8 false colour")
```







##########################
#JOIN FIELD DATA WITH COORDINATES FOR RAPIDEYE
##########################
```{r, warning=FALSE}
# merge Plot.data with plots (coordinates only)
data <- merge(Plot.data, plots[c("plot.id","lat","lon")], by="plot.id", all.x=TRUE)

# reproject coordinates from Lat/Lon to UTM
# add columns for "utm_x" and "utm_y"
data[c("utm_x","utm_y")] <- as.data.frame(project(cbind(data$lon, data$lat),
                                                  "+proj=utm +zone=30 ellps=WGS84"))

# plot image with plots
plotRGB(rapideye.30m, r=4, g=3, b=2, stretch="lin", main="RE with plots")
points(data$utm_x, data$utm_y, pch=19, col="yellow")
```



##########################
#JOIN FIELD DATA WITH COORDINATES FOR LANDSAT8
##########################
```{r, warning=FALSE}
# merge Plot.data with plots (coordinates only)
data <- merge(Plot.data, plots[c("plot.id","lat","lon")], by="plot.id", all.x=TRUE)

# reproject coordinates from Lat/Lon to UTM
# add columns for "utm_x" and "utm_y"
data[c("utm_x","utm_y")] <- as.data.frame(project(cbind(data$lon, data$lat),
                                                  "+proj=utm +zone=30 ellps=WGS84"))

# plot image with plots
plotRGB(landsat, r=4, g=3, b=2, stretch="lin", main="LS 8 with plots")
points(data$utm_x, data$utm_y, pch=19, col="yellow")
```




##########################
#EXTRACT PIXEL-VALUES FROM RAPIDEYE IMAGE
##########################

```{r, warning=FALSE}
# for further analysis, we need to extract pixel values for the plots
data[bands_re] <- as.data.frame(extract(rapideye.30m, cbind(data$utm_x, data$utm_y),
                                        method='bilinear'))

# see how data looks now
head(data)
```




##########################
#EXTRACT PIXEL-VALUES FROM LANDSAT8 IMAGE
##########################
```{r, warning=FALSE}
# for further analysis, we need to extract pixel values for the plots
data[bands_l8] <- as.data.frame(extract(landsat, cbind(data$utm_x, data$utm_y),
                                        method='bilinear'))
# see how data looks now
head(data)
```

##########################
#ACCURACY ASSESSMENT BY CROSS-VALIDATION FOR RAPIDEYE
##########################


```{r, warning=FALSE, message=FALSE}

# test accuracy for several variables
# make vector of y-variables
yvars <- c("ba.ha","agb.ha","h.mean","c.can")
k <- 3
method <- "msn"
xvars <- bands_re
par(mfrow=c(2,2)) #four variables are plotted in four panels
res_re <- c()
for(i in 1:length(yvars)){
  res_re <- rbind(res_re, cv.plot(data, yvars[i], xvars, k, method)) 
}
```



##########################
#ACCURACY ASSESSMENT BY CROSS-VALIDATION FOR LANDSAT8
##########################
```{r, warning=FALSE, message=FALSE}

# test accuracy for several variables
# make vector of y-variables
yvars <- c("ba.ha","agb.ha","h.mean","c.can")
k <- 3
method <- "msn"
xvars <- bands_l8
par(mfrow=c(2,2)) #four variables are plotted in four panels
res_l8 <- c()
for(i in 1:length(yvars)){
  res_l8 <- rbind(res_l8, cv.plot(data, yvars[i], xvars, k, method)) 
}
```


From the below, it can be seen that the landsat 8 is more accurate than the rapideye as its the RMSE values are lower which shows that the residuals (i.e the difference between the measured/observed values and the predictions) are lower. Although, the differences are not so much, it can be deduced that the landsat 8 provide more accurate predictions for the Basal area, the above-ground biomass, mean height of trees and the canopy cover. Hence, it would be more appropriate.

```{r, warning=FALSE, message=FALSE}
# organized results for rapideye
res_re <- cbind(yvars, res_re)
res_re <- as.data.frame(res_re)
names(res_re) <- c("Attribute","Bias","Bias (%)","RMSE","RMSE (%)","R.squared")
res_re

# organized results for landsat8
res_l8 <- cbind(yvars, res_l8)
res_l8 <- as.data.frame(res_l8)
names(res_l8) <- c("Attribute","Bias","Bias (%)","RMSE","RMSE (%)","R.squared")
res_l8
```


##########################
#PREDICT ATTRIBUTES TO PIXELS FOR RAPIDEYE
##########################
```{r, warning=FALSE, message=FALSE}
# create yai-object
# select vegetation attributes
yvars <- c("agb.ha","c.can")
xvars <- bands_re
y <- as.data.frame(data[,yvars])
x <- as.data.frame(data[,xvars])
k = 3
method = "msn"
msn <- yai(x=x, y=y, k=k, data=data, method=method)

# convert RapidEye image to data frame
rapideye.30m.df <- as.data.frame(rapideye.30m)

# add 160 observations to data frame
r160 <- as.data.frame(matrix(nrow=160, ncol=5, 0))
names(r160) <- bands_re
rapideye.30m.df <- rbind(r160, rapideye.30m.df)

# predict values for pixels
pred <- predict(msn, newdata=rapideye.30m.df, k=3, method="dstWeighted", observed=FALSE)
# ignore warning: this is why 160 rows were added above 


#RapidEye
# convert data frame back to raster attribute by attribute
# nrow and ncol depend on the image size
# set extent and projection based on the image
# aboveground biomass
agb_ha <- raster(matrix(pred$agb.ha, nrow=585, ncol=620, byrow=TRUE))
extent(agb_ha) <- rapideye.30m
projection(agb_ha) <- projection(rapideye.30m)

# canopy cover
cc <- raster(matrix(pred$c.can, nrow=585, ncol=620, byrow=TRUE))
extent(cc) <- rapideye.30m
projection(cc) <- projection(rapideye.30m)

# plot rasters
par(mfrow=c(1,2))
plot(agb_ha, main="Aboveground biomass (Mg/ha)")
plot(cc, main="Canopy cover (%)")

# compute mean based on the raster
round(cellStats(agb_ha, stat='mean'), 2)
round(cellStats(cc, stat='mean'), 2)
# should be close to the field based estimate
```



##########################
#PREDICT ATTRIBUTES TO PIXELS FOR LANDSAT8
##########################
```{r,warning=FALSE, message=FALSE}
# create yai-object
# select vegetation attributes
yvars <- c("agb.ha","c.can")
xvars <- bands_l8
y <- as.data.frame(data[,yvars])
x <- as.data.frame(data[,xvars])
k = 3
method = "msn"
msn <- yai(x=x, y=y, k=k, data=data, method=method)

# convert Landsat image to data frame
landsat.df <- as.data.frame(landsat)

# add 160 observations to data frame
r160 <- as.data.frame(matrix(nrow=160, ncol=6, 0))
names(r160) <- bands_l8
landsat.df <- rbind(r160, landsat.df)

# predict values for pixels
pred <- predict(msn, newdata=landsat.df, k=3, method="dstWeighted", observed=FALSE)
# ignore warning: this is why 160 rows were added above 


#Lansaat
# convert data frame back to raster attribute by attribute
# nrow and ncol depend on the image size
# set extent and projection based on the image

# aboveground biomass
agb_ha <- raster(matrix(pred$agb.ha, nrow=585, ncol=620, byrow=TRUE))
extent(agb_ha) <- landsat
projection(agb_ha) <- projection(landsat)

# canopy cover
cc <- raster(matrix(pred$c.can, nrow=585, ncol=620, byrow=TRUE))
extent(cc) <- landsat
projection(cc) <- projection(landsat)

# plot rasters
par(mfrow=c(1,2))
plot(agb_ha, main="Aboveground biomass (Mg/ha)")
plot(cc, main="Canopy cover (%)")

# compute mean based on the raster
round(cellStats(agb_ha, stat='mean'), 2)
round(cellStats(cc, stat='mean'), 2)
# should be close to the field based estimate
```


As it can be seen, the rapideye slightly underestimated the mean aboveground biomass and canopy cover. As mentioned earlier ,the rapideye has higher residual values and RMSE which reflects in the underestimation. Therefore, overall, the landsat 8 has better accuracy than the rapideye.